Scalar Feedback based Joint Time-Frequency Precoder Interpolation for MIMO-OFDM Systems

### 2020-03-19 paper/response_letter.tex

Thank you for this comment. We have fixed all instances of this error.

\textbf{2. `` The quantization approach described above... '' Actually,
  parameterization is described without presenting the quantizing
  method.}

\textbf{Response:} We thank the reviewer for pointing out this
error. We have reworded Section III to clarify the parameterization and quantization
in a suitable order. In addition, we have altered the text to include this:

\emph{The distribution of $\theta_{k,l}$ is used for
quantizing $\bf{\Theta}$ during initialization to reduce the
quantization error.  If $n$ bits are allocated to quantize a
parameter for the first time, the range of values that each parameter
takes is divided into $2^n$ optimal intervals using Lloyd's
quantization [16].}

\textbf{3. ``While tracking the $\phi_i$ parameters, is must be noted...''}

\textbf{Response:} It has been corrected to \emph{``While tracking the $\phi_{k,l}$ parameters, it must be noted...''}.

\textbf{4. ``we will use precoder parameters from both the past and future and
time instances to do the interpolation...''}

\textbf{Response:} Thank you. The typo has now been corrected to read
``we will use precoder parameters from both the past and future time instances to do the interpolation...''

\textbf{5. ``the parameters that determine the precoder is ...  = 10, ...''}

@@

``\emph{Independent time-frequency based approaches correspond to the
  method without joint-time frequency interpolation i.e. with
  prediction in time domain along with separate interpolation in the frequency
  domain.}''

\textbf{2. The quantization of $\phi_{k,l}$'s and $\theta_{k,l}$'s mentioned in Subsection III.A and the interpolation
process mentioned in Subsection III.C should be explained more clearly.}

\textbf{Response:}
Thanks for pointing this out. The explanation of the quantization
process has now been made more  clear by altering the text in III.A as
follows:

``\emph{If $n$ bits are allocated to quantize a parameter
for the first time, the range of values that each parameter takes is
divided into $2^n$ optimal intervals using Lloyd's
quantization [16].}''

### 2020-03-19 paper/main.tex
Add changes

e^{j\phi_{k,N_T }} \big)$ where  dimension $N_T \times N_T$. ${\bf 1}_{k-1}$ represents $k-1$ ones,

@@

The parameterization approach described above is related to the channel
state information feedback present in several wireless standards,
including 802.11ac~\cite{lou2013comparison}, although adaptive
quantization and feedback are not discussed. Further
sections show that the use of adaptive approaches can greatly enhance
performance with only very minor modifications to the system.
\subsection{Differential quantization and channel tracking}
\label{quantiz}
For channels that vary slowly with time, adaptive differential
quantization is an effective method for tracking parameters over time
using very few bits (often just one bit per parameter). For example,
while tracking a slowly varying discrete random process $x_n$ at time
$n$, the one bit can be used to determine the direction in which to
move in order to track the parameter. This can be denoted by
$\beta_{n} = \mbox{sign}(x_{n} - \hat{x}_{n-1})$, where $x_n$ is the
unquantized (accurate) value and $\hat{x}_{n-1}$ is the quantized
value. Further, the step size for the move can be adapted based on the
speed of channel variation as well the accuracy needed. If $\Delta_n$
is the step size, we define our adaptive quantizer as
\begin{equation}
\hat{x}_{n} = \hat{x}_{n-1} + \beta_{n}\Delta_{n} \mbox{ where }
\label{delta_eqn}
\Delta_{n} = \begin{cases}
    M \Delta_{n-1}, & \text{if $\beta_{n} = \beta_{n-1}$}\\
    \Delta_{n-1}/M , & \text{if $\beta_{n} \neq \beta_{n-1}$}.
  \end{cases}
\end{equation}
where $M$ is the scaling factor used during the positive and negative
moves. We initialize $\Delta_1$ as $\Delta_1 = |x_{2}-\hat{x}_1|$. This is performed separately for each of the quantized
parameters.

@@

For the adaptive channel tracking scheme, the arrows in
Fig.~\ref{fig:adpm-fig} show how previous quantized values along with
the 1-bit enhancement are used to find the new quantized value. Since
we use a single bit adaptation, the amount of feedback required is 1
bit for each scalar parameter that is being tracked. On average, this
can be reduced by half if we transmit only $\bf{\Theta}$ values for
one-time instance and only $\bf{\Phi}$ values for the next time
instant in an alternating fashion (the untransmitted value is assumed
to be the same as that at the previous instant). In this case, the
average number of bits required to quantize a channel matrix will be
$N_{R}(2N_{T} - N_R )/2$ (from~\cite{4114278}). While tracking
the $\phi_{k,l}$ parameters, it must be noted they may change abruptly
between $-\pi$ and $\pi$ due to jumps of $2\pi$. This is avoided by
unwrapping the phases to facilitate continuous tracking (i.e., jumps
in phase of magnitude close to $2\pi$ have been eliminated).
\subsection{Joint Time Frequency Interpolation}
\label{interp}
After obtaining the $\beta$ values (using the single bit feedback) at
the transmitter, we can construct the precoder at subcarrier indices
$pk$ or $pk+q$, keeping with the same notation as the previous
section. To construct precoders for rest of the subcarriers, we will use
precoder parameters from both the ``past'' and ``future'' time
instances to do the interpolation (using backtracking to ensure
causality). When using future time instances, the implication is that the precoder
estimation is improved at a later time instant i.e. after receiving
the feedback of the next set of subcarriers. Note that the precoder
used for transmission is causally obtained, while the ``future''
correction of past precoders merely improves the prediction
accuracy. In Fig.~\ref{fig:adpm-fig}, each parameter is interpolated
using the neighbouring subcarriers using bilinear interpolation
(equivalent to linear interpolation in two dimensions). These
interpolated values are then used to reconstruct the precoding
matrices all together by applying the reverse transformation of
breaking the precoder into independent parameters as given in
Equation~\ref{eq:decompose}. Thus, the approach here is able to
exploit the time and frequency correlation jointly to enhance
performance.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model
for the COST207 channel power delay
profiles~\cite{molisch2006cost259}. The BER achieved by
the proposed predictive quantization based precoder is compared
against the ideal precoder for 100 channel evolutions in each run and
averaged over 10 channel realizations (translating to averaging over $250,000$ bits). The error properties obtained by simulating transmission over these varying channels is sufficient to obtain good average performance properties. Simulations are performed for
normalised Doppler values of $f_dT_s = 3.41\times 10^{-6}$ (corresponding
to a velocity of 37.2 km/h for $f_c = 1$ GHz) with 1024 subcarriers, and with
$f_dT_s = 1.56 \times 10^{-5}$ (velocity of 172.8 km/h) with 64
subcarriers. The sampling time period $T_s$ was $5\times10^{-8}$
s. For both the situations, we considered $N_T=4$ transmit antennas
and and $N_R=2$ receive antennas. Therefore, using the approach
outlined in~\cite{4114278}, the parameters that determine the precoder is
$N_{R}(2N_{T} - N_R) = 12$, with five $\theta_{k,l}$s and and seven
$\phi_{k,l}$s. We assume that the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the slow and fast channel profiles
respectively (i.e. for the slower channels with 1024 subcarriers, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the faster channel, with 64 subcarriers, $0, 9, 18, \ldots 63$ would
be fed back).

>>>>

e^{j\phi_{k,N_T }} \big)$ with  dimension $N_T \times N_T$. ${\bf 1}_{k-1}$ represents $k-1$ ones,

@@

The parameterization approach described above is related to the channel
state information feedback present in several wireless standards,
including 802.11ac~\cite{lou2013comparison}, although adaptive
quantization and feedback are not discussed. Further
sections show that the use of adaptive approaches can greatly enhance
performance with only very minor modifications to the system.

\vspace{-0.5\baselineskip}
\subsection{Differential quantization and channel tracking}
\label{quantiz}
For channels that vary slowly with time, adaptive differential
quantization is an effective method for tracking parameters over time
using very few bits (often just one bit per parameter). For example,
while tracking a slowly varying discrete random process $x_n$ at time
$n$, the one bit can be used to determine the direction in which to
move in order to track the parameter. This can be denoted by
$\beta_{n} = \mbox{sign}(x_{n} - \hat{x}_{n-1})$, where $x_n$ is the
unquantized (accurate) value and $\hat{x}_{n-1}$ is the quantized
value. Further, the step size for the move can be adapted based on the
speed of channel variation as well the accuracy needed. If $\Delta_n$
is the step size, we define our adaptive quantizer as
\begin{equation}
\hat{x}_{n} = \hat{x}_{n-1} + \beta_{n}\Delta_{n} \mbox{ where }
\label{delta_eqn}
\Delta_{n} = \begin{cases}
    M \Delta_{n-1}, & \text{if $\beta_{n} = \beta_{n-1}$}\\
    \Delta_{n-1}/M , & \text{if $\beta_{n} \neq \beta_{n-1}$}.
  \end{cases}
\end{equation}
where $M$ is the scaling factor used during the positive and negative
moves. We initialize $\Delta_1$ as $\Delta_1 = |x_{2}-\hat{x}_1|$. This is performed separately for each of the quantized
parameters.

@@

For the adaptive channel tracking scheme, the arrows in
Fig.~\ref{fig:adpm-fig} show how previous quantized values along with
the 1-bit enhancement are used to find the new quantized value. Since
we use a single bit adaptation, the amount of feedback required is 1
bit for each scalar parameter that is being tracked. On average, this
can be reduced by half if we transmit only $\bf{\Theta}$ values for
one-time instance and only $\bf{\Phi}$ values for the next time
instant in an alternating fashion (the untransmitted value is assumed
to be the same as that at the previous instant). In this case, the
average number of bits required to quantize a channel matrix will be
$N_{R}(2N_{T} - N_R )/2$ (from~\cite{4114278}). While tracking
the $\phi_{k,l}$ parameters, it must be noted they may change abruptly
between $-\pi$ and $\pi$ due to jumps of $2\pi$. This is avoided by
unwrapping the phases to facilitate continuous tracking (i.e., jumps
in phase of magnitude close to $2\pi$ have been eliminated).

\vspace{-1\baselineskip}
\subsection{Joint Time Frequency Interpolation}
\label{interp}
After obtaining the $\beta$ values (using the single bit feedback) at
the transmitter, we can construct the precoder at subcarrier indices
$pk$ or $pk+q$, keeping with the same notation as the previous
section. To construct precoders for rest of the subcarriers, we will use
precoder parameters from both the ``past'' and ``future'' time
instances to do the interpolation (using backtracking to ensure
causality). When using future time instances, the implication is that the precoder
estimation is improved at a later time instant i.e. after receiving
the feedback of the next set of subcarriers. Note that the precoder
used for transmission is causally obtained, while the ``future''
correction of past precoders merely improves the prediction
accuracy. In Fig.~\ref{fig:adpm-fig}, each parameter is interpolated
using the neighbouring subcarriers using bilinear interpolation
(equivalent to linear interpolation in two dimensions). These
interpolated values are then used to reconstruct the precoding
matrices all together by applying the reverse transformation of
breaking the precoder into independent parameters as given in
Equation~\ref{eq:decompose}. Thus, the approach here is able to
exploit the time and frequency correlation jointly to enhance
performance.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model
for the COST207 channel power delay
profiles~\cite{molisch2006cost259}. The BER achieved by
the proposed predictive quantization based precoder is compared
against the ideal precoder for 100 channel evolutions in each run and
averaged over 10 channel realizations (translating to averaging over $125,000$ bits for 64-point DFT and
$2\times 10^6$ bits for the 1024-point DFT case). 
The error properties obtained by simulating transmission over these 
varying channels is sufficient to obtain good average performance properties. 
Simulations are performed for
normalised Doppler values of $f_dT_s = 3.41\times 10^{-6}$ (corresponding
to a velocity of 37.2 km/h for $f_c = 1$ GHz) with 1024 subcarriers, and with
$f_dT_s = 1.56 \times 10^{-5}$ (velocity of 172.8 km/h) with 64
subcarriers. The sampling time period $T_s$ was $5\times10^{-8}$
s. For both the situations, we considered $N_T=4$ transmit antennas
and and $N_R=2$ receive antennas. Therefore, using the approach
outlined in~\cite{4114278}, the parameters that determine the precoder is
$N_{R}(2N_{T} - N_R) = 12$, with five $\theta_{k,l}$s and and seven
$\phi_{k,l}$s. We assume that the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the slow and fast channel profiles
respectively (i.e. for the slower channels with 1024 subcarriers, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the faster channel, with 64 subcarriers, $0, 9, 18, \ldots 63$ would
be fed back).
